# Descriptive statistics and exploratory data analysis

## Measures of location

### Mean, median, and trimmed mean

For observations \(x_1,\ldots,x_n\), the arithmetic mean is

\[
\bar{x}=\frac{1}{n}\sum_{i=1}^{n}x_i.
\]

It uses every observation and is efficient for many regular distributions, but it can be strongly influenced by a long tail or extreme values. The median is the middle ordered value and is more resistant to outliers. A trimmed mean compromises between the two by removing a fixed proportion from each tail before averaging.

#### Arithmetic and robust location summaries

```{r location-examples}
x <- dat$crp_baseline_mg_L
location <- tibble(mean=mean(x, na.rm=TRUE),
                   median=median(x, na.rm=TRUE),
                   trimmed_mean_10pct=mean(x, trim=.10, na.rm=TRUE),
                   q25=quantile(x,.25,na.rm=TRUE), q75=quantile(x,.75,na.rm=TRUE));

kbl(location, digits=2, caption="Location summaries for baseline CRP")
```

### Weighted means

When observations carry legitimate representation or precision weights, the weighted mean is

\[
\bar{x}_w=\frac{\sum_i w_i x_i}{\sum_i w_i}.
\]

The workbook contains `sampling_weight` specifically for sampling demonstrations.However, the trial's primary treatment comparisons are **not** weighted by this educational variable.

#### Weighted mean for systolic blood pressure

```{r weighted-mean}
ok <- complete.cases(dat$sbp_baseline_mmHg, dat$sampling_weight)

tibble(unweighted=mean(dat$sbp_baseline_mmHg[ok]),
       weighted=weighted.mean(dat$sbp_baseline_mmHg[ok], dat$sampling_weight[ok])) |>
kbl(digits=2, caption="Unweighted and weighted baseline SBP")
```

## Measures of variability

### SD, IQR, and MAD

The sample variance measures the average squared spread of observations around the sample mean.

\[
s^2=\frac{\sum_{i=1}^{n}(x_i-\bar{x})^2}{n-1},
\]

and the sample standard deviation is

\[
s=\sqrt{s^2},
\]

which returns dispersion to the original measurement scale.

The interquartile range is

\[
IQR=Q_3-Q_1,
\]

the width of the middle 50% of observations.

The median absolute deviation (MAD) measures the typical distance of observations from the median.

\[
MAD=\operatorname{median}_{i}\left|x_i-\tilde{x}\right|,
\]

where \(\tilde{x}\) is the sample median. MAD is a robust measure of dispersion because it is based on absolute deviations from the median and is less sensitive to extreme observations than the standard deviation.

#### Measures of spread and dispersion

```{r variability-examples}
tibble(
  variable = c("Baseline SBP", "Baseline CRP"),
  SD = c(
    sd(dat$sbp_baseline_mmHg, na.rm=TRUE),
    sd(dat$crp_baseline_mg_L, na.rm=TRUE)
  ),
  IQR = c(
    IQR(dat$sbp_baseline_mmHg, na.rm=TRUE),
    IQR(dat$crp_baseline_mg_L, na.rm=TRUE)
  ),
  MAD_scaled = c(
    mad(dat$sbp_baseline_mmHg, na.rm=TRUE),
    mad(dat$crp_baseline_mg_L, na.rm=TRUE)
  ),
  MAD_raw = c(
    mad(dat$sbp_baseline_mmHg, constant=1, na.rm=TRUE),
    mad(dat$crp_baseline_mg_L, constant=1, na.rm=TRUE)
  )
) |>
  kbl(digits=2, caption="Classical and robust measures of spread")
```

### Raw versus scaled MAD

The distinction is important. The lecture formula is the **raw** median absolute deviation. R's `mad()` multiplies that quantity by a consistency constant by default; `mad(x, constant=1)` reproduces the raw formula.

## Percentiles and distributional summaries

A percentile is a cut point in the ordered data: the \(p\)th quantile is a value at or below which approximately \(100p\%\) of observations fall. Percentiles are especially useful for skewed biomarkers and for identifying clinically relevant distribution tails.

### Quantiles in R

```{r percentiles}
quantile(dat$crp_baseline_mg_L, probs=c(.05,.10,.25,.50,.75,.90,.95), na.rm=TRUE)
```

## Grouped descriptive statistics

A good descriptive table reports the number of non-missing observations and summaries appropriate to the distribution. SBP is reasonably summarized by mean and SD; CRP is long-tailed, so median and IQR are more informative.

### Summary by randomized arm

```{r grouped-summary}
grouped <- dat |>
  group_by(treatment_arm) |>
  summarise(
    n=n(),
    across(c(age_years, sbp_baseline_mmHg),
           list(mean=~mean(.x,na.rm=TRUE), sd=~sd(.x,na.rm=TRUE)),
           .names="{.fn}_{.col}"),
    across(c(crp_baseline_mg_L, biomarker_baseline_ng_mL),
           list(median=~median(.x,na.rm=TRUE), IQR=~IQR(.x,na.rm=TRUE)),
           .names="{.fn}_{.col}"),
    .groups="drop"
  )

kbl(grouped, digits=2, caption="Selected baseline summaries by randomized arm")
```

In a randomized trial, baseline tables **describe the sample**. Routine p-values are usually unnecessary because baseline differences arise by chance from randomization.


## Categorical data

For categorical variables, frequencies and proportions are the natural summaries.

### Frequencies and proportions

```{r categorical-summary}
severity <- dat |> count(treatment_arm, baseline_severity, name="n") |>
                  group_by(treatment_arm) |>
                  mutate(percent=100*n/sum(n)) |> ungroup()

kbl(severity, digits=1, caption="Baseline severity by treatment arm")
```

#### Clinical-response proportions

```{r response-table}
response_tab <- table(dat$treatment_arm, dat$response);
response_tab;
prop.table(response_tab, margin=1)
```

Row proportions provide an initial indication of **association between two categorical variables**: differences across treatment arms may suggest an association between treatment type and clinical response.  
The corresponding inferential analysis is developed in Chapters 3 and 6.

## Histograms, density plots, and boxplots

A histogram is a visual **frequency** table; its appearance depends on bin width.  
A kernel density plot smooths the empirical distribution and uses **density**, not raw count or a direct category proportion, on the vertical axis. Its scale is defined so that the total area under the density curve is 1.  
A boxplot compactly displays the median, quartiles, and points beyond the conventional whiskers. 

### Distribution shape

```{r baseline-distributions, fig.height=7.0, fig.cap="selected baseline distributions by randomized treatment arm."}
base_long <- dat |>
        dplyr::select(treatment_arm, sbp_baseline_mmHg, crp_baseline_mg_L, biomarker_baseline_ng_mL, qol_baseline_0_100) |>
        pivot_longer(-treatment_arm, names_to="measure", values_to="value")

ggplot(base_long, aes(value, fill=treatment_arm)) +
  geom_histogram(bins=28, alpha=.55, position="identity", na.rm=TRUE) +
  facet_wrap(~measure, scales="free", ncol=2) +
  scale_fill_manual(values=pal) +
  labs(x=NULL, y="Count", fill="Treatment")
```

### Groupwise boxplot

```{r crp-boxplot, fig.cap="Baseline CRP by treatment arm."}
ggplot(dat, aes(treatment_arm, crp_baseline_mg_L, fill=treatment_arm)) +
  geom_boxplot(width=.6, outlier.alpha=.45, na.rm=TRUE) +
  scale_fill_manual(values=pal, guide="none") +
  labs(x=NULL, y="Baseline CRP (mg/L)")
```

## Outliers: observations to understand, not delete automatically

A boxplot point beyond 1.5 IQR is not automatically an erroneous observation/ outlier. Before exclusion, ask whether the value is impossible, caused by a measurement or entry error, inconsistent with source data, or simply a legitimate extreme value.  
Deleting a genuine observation because it weakens significance is unacceptable.  
Robust summaries, transformations, and sensitivity analyses are preferable when extreme but valid observations materially affect a model.

## EDA checklist

Before inferential analysis, the analyst should be able to state:
- the observational unit and whether observations are independent, paired, clustered, or repeated;
- the outcome type and clinically meaningful direction of effect;
- treatment/reference categories and factor coding;
- the amount and pattern of missingness;
- plausible ranges and units;
- distributional shape, outliers, and group-specific spread;
- whether a transformation has a scientific interpretation;
- which baseline variables are genuinely pre-treatment covariates; 
- which variables are post-randomization and therefore require caution in causal treatment models.

For example, the adherence to treatment percentage`adherence_pct`is measured after randomization and may itself be affected by treatment. Adjusting for it can therefore change the treatment-effect estimand and introduce bias. For this reason, it is excluded from the primary treatment-effect models and considered only in clearly labeled associational analyses.

