# Sampling distributions and confidence intervals

Statistical inference treats the observed sample as one outcome of a repeatable sampling process. A key distinction is between the **distribution of individual observations** and the **sampling distribution of a statistic**. For example, CRP measurements may be strongly right-skewed while the distribution of sample means is much closer to normal.

## Random and stratified sampling

A **simple random sample** gives each eligible observation an equal chance of selection. **Stratified sampling** divides the population into predefined groups and samples within each group. If sampling fractions differ across strata, population estimates may require sampling weights.

```{r sampling-examples}
simple_60 <- dat |> slice_sample(n=60)
stratified_60 <- dat |> group_by(treatment_arm) |> slice_sample(n=20) |> ungroup()

table(simple_60$treatment_arm)
table(stratified_60$treatment_arm)
```

## Sampling bias

Bias occurs when the sampling process systematically produces a sample that does not represent the target population. For example, sampling only older participants may give distorted estimates if the target population includes all ages.

```{r sampling-bias}
tibble(
  sample=c("All participants","Age ≥65 years"),
  mean_age=c(mean(dat$age_years), mean(dat$age_years[dat$age_years>=65])),
  mean_sbp=c(mean(dat$sbp_baseline_mmHg), mean(dat$sbp_baseline_mmHg[dat$age_years>=65]))
) |>
  kbl(digits=2, caption="A deliberately biased sampling example")
```

## Sampling distribution and standard error

The **sampling distribution** describes how a statistic would vary across repeated samples. Its standard deviation is the **standard error (SE)**.

For the sample mean,

\[
SE(\bar X)=\frac{\sigma}{\sqrt n}.
\]

Because the population standard deviation \(\sigma\) is usually unknown, the SE is estimated using the sample standard deviation:

\[
\widehat{SE}(\bar X)=\frac{s}{\sqrt n}.
\]

### Repeated-sampling simulation

The observed baseline SBP values are treated as an empirical population, and samples of 40 observations are repeatedly drawn with replacement.

```{r sampling-distribution}
sbp <- na.omit(dat$sbp_baseline_mmHg)
means40 <- replicate(3000, mean(sample(sbp, 40, replace=TRUE)))

tibble(
  empirical_mean=mean(means40),
  empirical_SE=sd(means40),
  theoretical_SE=sd(sbp)/sqrt(40)
) |>
  kbl(digits=3, caption="Sampling distribution of the mean")
```

```{r sampling-distribution-plot, fig.cap="Simulated sampling distribution of mean baseline SBP for samples of size 40."}
ggplot(tibble(mean_sbp=means40), aes(mean_sbp)) +
  geom_histogram(bins=35, fill="#547D8E", colour="white") +
  labs(x="Sample mean SBP (mmHg)", y="Simulation count")
```

## Central limit theorem

The **central limit theorem (CLT)** explains why the sampling distribution of a mean becomes approximately normal as sample size increases, even when the original observations are skewed. It does not imply that the original data become normal.

```{r clt-simulation}
crp <- na.omit(dat$crp_baseline_mg_L)

clt <- bind_rows(
  tibble(value=replicate(2000, mean(sample(crp,5,TRUE))), n="n = 5"),
  tibble(value=replicate(2000, mean(sample(crp,30,TRUE))), n="n = 30"),
  tibble(value=replicate(2000, mean(sample(crp,100,TRUE))), n="n = 100")
)
```

```{r clt-plot, fig.height=4.5, fig.cap="Sampling distributions of the mean for a right-skewed CRP variable."}
ggplot(clt, aes(value)) +
  geom_histogram(bins=32, fill="#315B74", colour="white") +
  facet_wrap(~n, scales="free") +
  labs(x="Sample mean CRP (mg/L)", y="Simulation count")
```

As \(n\) increases, the sampling distribution becomes narrower and more symmetric.

## Confidence intervals

A confidence interval gives a range of plausible values for the population mean. When the population variance is unknown, the sample standard deviation \(s\) is used and the interval is based on the \(t\)-distribution.

\[
\bar{x}\pm t_{1-\alpha/2,\,n-1}\frac{s}{\sqrt n}.
\]

The critical value t is the multiplier taken from the \(t\)-distribution; It determines how many standard errors are used in the confidence interval, and it depends on the sample size.

```{r mean-ci}
t.test(dat$sbp_baseline_mmHg, conf.level=.95)$conf.int
```

### Interpretation

A 95% confidence interval is produced by a method that would contain the true parameter in about 95% of repeated samples, assuming the model assumptions hold.

For a two-sided test at \(\alpha=0.05\), a 95% confidence interval that excludes the null value usually corresponds to \(p<0.05\). The interval also shows the estimated effect and the uncertainty around it [@greenland2016; @wasserstein2016].