# Bootstrap and probability models

## Bootstrap inference

### Bootstrap algorithm

The nonparametric bootstrap approximates a sampling distribution by repeatedly sampling **with replacement** from the observed sample and recalculating the statistic. 

For a statistic \(\hat\theta\), bootstrap replicates \(\hat\theta_1^*,\ldots,\hat\theta_B^*\) give an estimated standard error

\[
\widehat{SE}_{boot}=SD(\hat\theta_1^*,\ldots,\hat\theta_B^*).
\]

#### Bootstrap distribution of the median

```{r bootstrap-median}
B <- 5000;
crp_med <- median(crp);
boot_med <- replicate(B, median(sample(crp, replace=TRUE)))
boot_ci <- quantile(boot_med, c(.025,.975));

tibble(estimate=crp_med, bootstrap_SE=sd(boot_med), lower=boot_ci[1], upper=boot_ci[2]) |>
kbl(digits=3, caption="Bootstrap percentile interval for the baseline CRP median")
```

#### Bootstrap uncertainty visualization

```{r bootstrap-median-plot, fig.cap="Bootstrap distribution of the baseline CRP median."}
ggplot(tibble(estimate=boot_med), aes(estimate)) +
  geom_histogram(bins=35, fill="#9A3E76", colour="white") +
  geom_vline(xintercept=crp_med, linetype=2) +
  labs(x="Bootstrap median CRP (mg/L)", y="Count")
```

### Design-respecting resampling

Resampling should preserve the original study design. Paired observations are resampled together, clustered data are resampled by cluster, and randomized trials should follow the original randomization scheme. Otherwise, uncertainty estimates may be misleading.


### What bootstrap cannot repair

The bootstrap is **not** a remedy for an extremely small or unrepresentative sample. Resampling can approximate the empirical sampling distribution only from information already present in the observed data; it cannot manufacture unobserved tails, subgroups, or biological heterogeneity [@efron1993].

## Standardization and normal quantiles

### Z-score standardization

A z-score expresses a value in standard-deviation units:

\[
z_i=\frac{x_i-\bar x}{s}.
\]

```{r standardization}
sbp_z <- as.numeric(scale(dat$sbp_baseline_mmHg));
summary(sbp_z);

c(Phi_1_96=pnorm(1.96), z_975=qnorm(.975))
```

### Q–Q plot assessment of distributional shape

A normal Q–Q plot compares empirical quantiles with theoretical normal quantiles. Departures in the center, tails, and asymmetry are more informative than a single normality-test p-value.

```{r qq-comparison, fig.height=4.2, fig.cap="Q–Q plots contrast approximately symmetric SBP with strongly right-skewed CRP."}
qq_dat <- dat |>
  dplyr::select(sbp_baseline_mmHg, crp_baseline_mg_L) |>
  pivot_longer(everything(), names_to="variable", values_to="value")

ggplot(qq_dat, aes(sample=value)) +
  stat_qq(alpha=.45, na.rm=TRUE) +
  stat_qq_line(na.rm=TRUE) + facet_wrap(~variable, scales="free") +
  labs(x="Theoretical normal quantiles", y="Observed quantiles")
```

## Student t distribution and degrees of freedom

When the population standard deviation is unknown and estimated from the sample, the standardized mean follows Student’s \(t\) distribution.  
It has heavier tails than the normal distribution, especially for small samples, reflecting the extra uncertainty from estimating the standard deviation.

### Student t distribution and degrees of freedom
```{r t-distribution}
c(
  t_975_df10 = qt(p = 0.975, df = 10),
  t_975_df50 = qt(p = 0.975, df = 50),
  normal_975 = qnorm(p = 0.975),
  P_T20_le_2 = pt(q = 2, df = 20)
)
```

Here, `p` is the cumulative probability, `df` is degrees of freedom, and `q` is the t-value at which the cumulative probability is evaluated.

## Binomial distribution

The binomial distribution models the number of successes in \(n\) independent trials when each trial has the same probability of success, \(p\).

\[
X\sim\operatorname{Binomial}(n,p),\qquad E(X)=np,\quad Var(X)=np(1-p).
\]

#### Binomial probability model

```{r binomial-example}
c(
  P_exactly_20 = dbinom(x=20, size=50, prob=.40),
  P_at_most_20 = pbinom(q=20, size=50, prob=.40)
)
```

Here, `dbinom()` gives the probability of exactly 20 successes, whereas `pbinom()` gives the cumulative probability of 20 or fewer successes.

The binomial distribution is widely used for binary outcomes and forms the basis of binomial regression models.


## Poisson distribution

### Poisson counts

The Poisson distribution models the number of events occurring in a fixed time, area, or other interval when events occur independently at an approximately constant rate.

\[
P(Y=y)=\frac{e^{-\lambda}\lambda^y}{y!},\quad y=0,1,2,\ldots
\]

#### Poisson count model

```{r poisson-example}
lambda <- mean(dat$ae_count_12w, na.rm=TRUE)

c(
  mean_count = lambda,
  P_exactly_2 = dpois(x=2, lambda=lambda),
  P_at_most_2 = ppois(q=2, lambda=lambda)
)
```
`x=2` in `dpois()` asks for the probability of **exactly 2 events**, \(P(Y=2)\).
`q=2` in `ppois()` asks for the **cumulative probability of 2 or fewer events**, \(P(Y\le2)\).
`lambda` is the **expected or average number of events**.

## Parametric and semiparametric survival models

The exponential and Weibull distributions model **time until an event occurs**.  
The exponential model assumes a constant hazard over time, whereas the Weibull model allows the hazard to increase or decrease over time.

The **Cox proportional hazards model** is different: it does not assume a specific survival-time distribution. Instead, it models how predictors affect the hazard of an event. It assumes that the **hazard ratio between groups remains approximately constant over time**.

### Exponential and Weibull survival probabilities

```{r survival-distributions}
c(
  exponential = 1 - pexp(q=30, rate=1/90),
  weibull = 1 - pweibull(q=30, shape=1.4, scale=100)
)
```
These values represent the probability of remaining event-free beyond 30 days, \(P(T>30)\).

## Bootstrap and permutation distributions

Bootstrap and permutation methods both use repeated resampling, but they answer different questions.

- **Bootstrap:** estimates the sampling distribution of a statistic by resampling observations **with replacement**.
- **Permutation:** estimates the distribution of a test statistic under the **null hypothesis** by shuffling group labels.

Bootstrap is mainly used to quantify **uncertainty in an estimate**, whereas permutation is mainly used for **hypothesis testing**.

### Bootstrap example: mean baseline Systolic blood pressure

```{r bootstrap-example}
set.seed(123)
boot_mean <- replicate(2000, mean(sample(dat$sbp_baseline_mmHg, replace=TRUE), na.rm=TRUE))

tibble(
  observed_mean=mean(x),
  bootstrap_mean=mean(boot_mean),
  bootstrap_SE=sd(boot_mean),
  CI_lower=quantile(boot_mean,.025),
  CI_upper=quantile(boot_mean,.975)
)
```

Each bootstrap sample has the same size as the original sample and is drawn **with replacement**. Repeating this many times produces a bootstrap distribution of the mean. Its standard deviation estimates the standard error.

### Permutation example: difference in mean SBP change

```{r permutation-example}
d <- dat |> filter(treatment_arm %in% c("Control","Drug_A")) |>
            mutate(sbp_change=sbp_week12_mmHg-sbp_baseline_mmHg) |>
            drop_na(treatment_arm, sbp_change)

obs_diff <- with(d, mean(sbp_change[treatment_arm=="Drug_A"]) -
                    mean(sbp_change[treatment_arm=="Control"]))

set.seed(123)
perm_diff <- replicate(2000, {
  gp <- sample(d$treatment_arm)
  mean(d$sbp_change[gp=="Drug_A"]) - mean(d$sbp_change[gp=="Control"])
})

p_value <- (sum(abs(perm_diff)>=abs(obs_diff))+1)/(length(perm_diff)+1)
c(observed_difference=obs_diff, permutation_p_value=p_value)
```

In each permutation, the SBP changes remain fixed while the **treatment labels are shuffled**. This generates the distribution expected under the null hypothesis.

The **permutation p-value** is the proportion of permuted differences that are at least as extreme as the observed difference:


\[
p=
\frac{\text{number of permuted statistics greater than or equal the observed statistic}}
{\text{number of permutations}}.
\]

A small p-value indicates that the observed group difference would be uncommon under the null hypothesis. 
