# Continuous two-group comparison

Continuous outcomes are commonly compared either between **different groups of participants** or between **linked measurements from the same participants**. These designs require different analyses because they contain different sources of variability.

For independent groups, inference concerns a contrast between populations, such as a difference in mean biomarker change between Control and Drug_A. For paired measurements, inference concerns the distribution of the within-person differences, such as the change in systolic blood pressure from baseline to week 12.

The design and estimand should be specified before examining assumptions. Distributional diagnostics then assess whether a mean-based model is reasonable and, for independent groups, whether a common-variance model is credible.

## Decision framework

| Design | Target | Method | Main consideration |
|---|---|---|---|
| independent groups | difference in means | Welch two-sample t-test | unequal variances allowed |
| independent groups | difference in means | pooled Student t-test | requires a common variance |
| independent groups | rank-based contrast | Wilcoxon–Mann–Whitney test | rank/distributional estimand |
| paired measurements | mean within-person difference | paired t-test | normality concerns the paired differences |
| paired measurements | rank-based paired contrast | Wilcoxon signed-rank test | interpretation depends on the distribution of differences |

Welch's test is generally the safer mean-based procedure for independent groups because it does not require equal variances. The pooled Student test remains useful when a common-variance model is justified.

# Independent two-group example: biomarker change

## Define the estimand

For Control and Drug_A, define biomarker change as

\[
\Delta B=B_{12}-B_0.
\]

The target is the difference between the groups in **mean biomarker change**.

## Descriptive statistics

```{r independent-biomarker-data}
ind_bio <- dat |> filter(treatment_arm %in% c("Control","Drug_A"), !is.na(biomarker_change)) |> droplevels()

ind_sum <- ind_bio |> group_by(treatment_arm) |>
                      summarise(
                      n=n(),
                      mean=mean(biomarker_change), sd=sd(biomarker_change),
                      median=median(biomarker_change), IQR=IQR(biomarker_change), .groups="drop")

kbl(ind_sum, digits=2, caption="Biomarker change in Control and Drug_A")
```

The mean and SD summarize the outcome on the scale relevant to mean comparisons. The median and IQR provide additional information about asymmetry and spread.

## Distribution and influential observations

```{r independent-biomarker-boxplot, fig.height=4.2, fig.cap="Distribution of biomarker change in Control and Drug_A."}
ggplot(ind_bio, aes(treatment_arm, biomarker_change)) +
  geom_boxplot(width=.55, outlier.shape=NA) +
  geom_jitter(width=.08, alpha=.30, size=1.3) +
  labs(x="Treatment group", y="Biomarker change")
```

The plot shows group location, spread, asymmetry, and individual observations. Large differences in spread are relevant to the pooled Student test, while extreme observations may affect any mean-based analysis.

## Normality assessment

Normality is assessed separately within each independent group.

### Q–Q plots

```{r independent-biomarker-qq, fig.height=4.0, fig.cap="Q–Q plots of biomarker change by treatment group."}
ggplot(ind_bio, aes(sample=biomarker_change)) +
  stat_qq(alpha=.45) + 
  stat_qq_line() + 
  facet_wrap(~treatment_arm) +
  labs(x="Theoretical normal quantiles", y="Observed quantiles")
```

Approximate alignment with the reference line supports a normal model. Systematic curvature suggests skewness, while marked tail departures may indicate heavy tails or influential observations.

### Shapiro–Wilk tests

```{r independent-biomarker-shapiro}
shapiro_bio <- ind_bio |> group_by(treatment_arm) |> group_modify(~{
  z <- shapiro.test(.x$biomarker_change)
  tibble(W=unname(z$statistic), p_value=z$p.value)
})

kbl(shapiro_bio, digits=3, caption="Shapiro–Wilk tests for biomarker change")
```

The Shapiro–Wilk test evaluates

\[
H_0:\text{the observations are compatible with a normal distribution}.
\]

A small Shapiro–Wilk p-value indicates departure from normality, but test choice should not depend on it alone. With a large sample and only mild Q–Q plot deviations, a t-test is usually still appropriate.
For independent groups, use **Welch's t-test** if variances differ.
If non-normality is substantial, especially with strong skewness or influential outliers, consider the **Wilcoxon–Mann–Whitney test**. For paired data, assess the **paired differences** and use either a paired t-test or Wilcoxon signed-rank test accordingly.

## Equality of variance

A median-centered Levene test, often called the **Brown–Forsythe test**, assesses whether the group spreads differ.

```{r independent-biomarker-variance}
lev_bio <- car::leveneTest(biomarker_change~treatment_arm, data=ind_bio, center=median)
lev_bio
```

The null hypothesis is

\[
H_0:\sigma_1^2=\sigma_2^2.
\]

Evidence of unequal variances argues against the pooled Student test. Welch's test does not require equal population variances.

## Automated test selection

The following block combines the diagnostic results and performs one selected analysis.

```{r independent-test-selection}
normal_ok <- all(shapiro_bio$p_value>=.05)
levene_p <- as.numeric(lev_bio[1,"Pr(>F)"])
equal_var <- levene_p>=.05

if(normal_ok && equal_var){
  selected_test <- "Pooled Student t-test"
  fit <- t.test(biomarker_change~treatment_arm, data=ind_bio, var.equal=TRUE)
  reason <- "Normality acceptable; no evidence of unequal variances."
} else if(normal_ok){
  selected_test <- "Welch two-sample t-test"
  fit <- t.test(biomarker_change~treatment_arm, data=ind_bio, var.equal=FALSE)
  reason <- "Normality acceptable; variances differ."
} else {
  selected_test <- "Wilcoxon–Mann–Whitney rank-sum test"
  fit <- wilcox.test(biomarker_change~treatment_arm, data=ind_bio, exact=FALSE, conf.int=TRUE)
  reason <- "Evidence of departure from normality; rank-based analysis selected."
}

tibble(selected_test, reason) |> kbl(caption="Automated test selection")
broom::tidy(fit) |> kbl(digits=3, caption="Selected independent-group analysis")
```

The operational rule is

| Normality | Variance assessment | Selected analysis |
|---|---|---|
| acceptable | similar | pooled Student t-test |
| acceptable | unequal | Welch t-test |
| substantial departure | either | Wilcoxon–Mann–Whitney test |

This is the general rule, however the threshold \(p=0.05\) should not be interpreted as a universal boundary. Q–Q plots, influential observations, sample size, and the intended estimand remain relevant.

## Mean-based tests

For the pooled Student test,

\[
t=
\frac{\bar y_1-\bar y_2}
{s_p\sqrt{1/n_1+1/n_2}},
\]

where \(s_p\) is the pooled estimate of the common SD.

Welch's test instead uses

\[
SE(\bar y_1-\bar y_2)
=
\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}},
\]

and estimates the degrees of freedom using the Welch–Satterthwaite approximation.

Both procedures estimate a **difference in means**. Their distinction concerns how uncertainty is estimated when group variances differ.

## Wilcoxon–Mann–Whitney test

The Wilcoxon rank-sum, or Mann–Whitney, procedure replaces the original measurements by their ranks and evaluates the relative ordering of observations from the two groups.

If \(R_1\) is the rank sum for group 1,

\[
U_1=R_1-\frac{n_1(n_1+1)}{2}.
\]

It should not automatically be described as a test of equal medians. A simple median or location-shift interpretation requires additional distributional conditions.

# Paired t-test and Wilcoxon signed-rank test

## Define the paired difference

For Drug_A participants, SBP is measured at baseline and again at week 12 in the **same participant**. The paired difference for participant \(i\) is

\[
d_i=\text{SBP}_{i,\text{week 12}}-\text{SBP}_{i,\text{baseline}}.
\]

Thus, \(d_i\) is the participant's SBP change. A negative value indicates a reduction in SBP.

```{r paired-sbp-data}
pair_sbp <- dat |> filter(treatment_arm=="Drug_A") |>
                  drop_na(sbp_baseline_mmHg, sbp_week12_mmHg) |>
                  mutate(sbp_change=sbp_week12_mmHg-sbp_baseline_mmHg)

pair_sbp |> summarise(
            n=n(),
            mean_baseline=mean(sbp_baseline_mmHg), mean_week12=mean(sbp_week12_mmHg),
            mean_change=mean(sbp_change), sd_change=sd(sbp_change), median_change=median(sbp_change)
          ) |> kbl(digits=2, caption="Drug_A paired SBP summary")
```

The paired analysis is based on `sbp_change`. The baseline and follow-up measurements are not treated as two independent samples.

## Participant-level changes

```{r paired-sbp-lines, fig.height=4.5, fig.cap="Participant-level SBP from baseline to week 12 in Drug_A."}
pair_long <- pair_sbp |>
            dplyr::select(patient_id, sbp_baseline_mmHg, sbp_week12_mmHg) |>
            tidyr::pivot_longer(-patient_id, names_to="time", values_to="sbp") |>
            mutate(time=factor(time, levels=c("sbp_baseline_mmHg","sbp_week12_mmHg"),
                               labels=c("Baseline","Week 12")))

ggplot(pair_long, aes(time, sbp, group=patient_id)) +
    geom_line(alpha=.18) + 
    geom_point(alpha=.35, size=1.2) +
    labs(x=NULL, y="SBP (mmHg)")
```

The paired-line plot shows both the direction and magnitude of participant-level changes.

## Distribution of paired differences

```{r paired-sbp-boxplot, fig.height=3.8, fig.cap="Distribution of within-participant SBP changes in Drug_A."}
ggplot(pair_sbp, aes(x="", y=sbp_change)) +
  geom_boxplot(width=.25, outlier.shape=NA) +
  geom_jitter(width=.08, alpha=.30, size=1.3) +
  labs(x=NULL, y="SBP change (mmHg)")
```

For paired data, analyze the **change within each participant**. Here, `sbp_change` is the difference between week-12 and baseline SBP. The boxplot is therefore used to check whether these individual changes are symmetric and whether any participants have unusually large increases or decreases.

## Normality of paired differences

### Q–Q plot

```{r paired-sbp-qq, fig.height=4.0, fig.cap="Q–Q plot of within-participant SBP changes in Drug_A."}
ggplot(pair_sbp, aes(sample=sbp_change)) +
  stat_qq(alpha=.5) +
  stat_qq_line() +
  labs(x="Theoretical normal quantiles", y="Observed change quantiles")
```

### Shapiro–Wilk test

```{r paired-sbp-shapiro}
shapiro_pair <- shapiro.test(pair_sbp$sbp_change)

tibble(W=unname(shapiro_pair$statistic), p_value=shapiro_pair$p.value) |>
  kbl(digits=3, caption="Shapiro–Wilk test for paired SBP differences")
```

The relevant normality assumption concerns the distribution of the **within-participant SBP changes**, \(d_i\).
Equal baseline and week-12 variances are not required, so Levene's test is not used for the paired t-test.

## Automated paired test selection

```{r paired-test-selection}
normal_diff <- shapiro_pair$p.value>=.05

if(normal_diff){
  selected_test <- "Paired t-test"
  fit <- t.test(pair_sbp$sbp_week12_mmHg, pair_sbp$sbp_baseline_mmHg, paired=TRUE)
  reason <- "Paired differences are reasonably compatible with normality."
} else {
  selected_test <- "Wilcoxon signed-rank test"
  fit <- wilcox.test(pair_sbp$sbp_week12_mmHg, pair_sbp$sbp_baseline_mmHg,
                     paired=TRUE, exact=FALSE, conf.int=TRUE)
  reason <- "Paired differences show evidence of departure from normality."
}

tibble(selected_test, reason) |> kbl(caption="Automated paired-test selection")
broom::tidy(fit) |> kbl(digits=3, caption="Selected paired analysis")
```

The operational rule is

| Distribution of paired differences | Selected analysis |
|---|---|
| approximately normal | paired t-test |
| substantial departure from normality | Wilcoxon signed-rank test |

For the paired t-test,

\[
t=\frac{\bar d}{s_d/\sqrt n},
\]

where \(\bar d\) is the mean within-person change and \(s_d\) is the SD of the paired differences. The estimate and confidence interval are expressed directly in mmHg.

The Wilcoxon signed-rank test compares paired differences using their ranks. It is an alternative to the paired t-test when the differences are not well suited to a mean-based analysis.

## Reporting the selected analysis

Report the **effect estimate, 95% confidence interval, and p-value**, together with the descriptive statistics that give the result context.

For independent groups, state the test used and the assumption that determined the choice. For paired data, report the **average within-participant change** and the corresponding paired-test result.


# Multiple testing

When several hypotheses are tested, each test creates another opportunity to obtain a small p-value by chance. Multiplicity adjustment accounts for this when the hypotheses belong to the same planned family of comparisons.

The following example compares Drug_A with Control for three continuous outcomes using Welch's two-sample t-test.

```{r cat-multiplicity}
outcomes <- c(SBP="sbp_change",
              Biomarker="biomarker_change",
              QoL="qol_change")

p_raw <- sapply(outcomes,\(v){
  d <- dat |> filter(treatment_arm %in% c("Control","Drug_A")) |> drop_na(all_of(v))
  t.test(d[[v]][d$treatment_arm=="Drug_A"],
         d[[v]][d$treatment_arm=="Control"],
         var.equal=FALSE)$p.value
})

tibble(
  outcome=names(outcomes),
  p_raw=p_raw,
  p_Holm=p.adjust(p_raw,method="holm"),
  p_BH=p.adjust(p_raw,method="BH")
) |> kbl(digits=4,
         caption="Raw and multiplicity-adjusted p-values for three outcomes")
```

## Holm and Benjamini–Hochberg adjustments

**Holm adjustment** controls the **family-wise error rate**, the probability of making at least one false-positive conclusion within the defined family of tests. It is appropriate when strong control of false-positive claims is important.

**Benjamini–Hochberg adjustment** controls the **false discovery rate**, the expected proportion of false positives among the results declared significant. It is commonly used when a larger set of hypotheses is examined.

| Goal | Adjustment |
|---|---|
| strongly limit the chance of any false-positive claim | **Holm** |
| control the proportion of false discoveries among significant results | **Benjamini–Hochberg** |

Multiplicity adjustment changes the p-values used for inference; it does **not** change the observed effect estimates.
