# Categorical Inference

Categorical outcomes are analyzed according to whether observations are **independent or paired**.

For independent groups, Pearson's chi-square test or Fisher's exact test assesses association between categorical variables. The choice between them depends mainly on whether the expected cell counts are large enough for the chi-square approximation. For binary outcomes measured twice in the same participants, the observations are paired and **McNemar's test** is used instead.

## Choosing the categorical test

| Data structure | Condition | Test |
|---|---|---|
| independent categorical observations | expected counts adequate | **Pearson chi-square test** |
| independent categorical observations | expected counts sparse | **Fisher's exact test** |
| paired binary observations | same participants measured twice | **McNemar's test** |

For independent observations, a practical criterion for using the Pearson chi-square approximation is based on the **expected cell counts, not the observed counts**: no expected count should be below 1, and no more than 20% of expected counts below 5.
If these conditions are not met, Fisher's exact test is preferred.

# Independent categorical outcomes

## Pearson chi-square and Fisher exact tests

Pearson's chi-square test compares the **observed counts**, \(O_{ij}\), with the counts expected if the two categorical variables were independent.

The expected count in cell \(i,j\) is

\[
E_{ij}=\frac{n_{i}n_{j}}{n},
\]

and the chi-square statistic is

\[
\chi^2=\sum_i\sum_j\frac{(O_{ij}-E_{ij})^2}{E_{ij}}.
\]

A larger discrepancy between observed and expected counts produces a larger chi-square statistic and stronger evidence that the variables are associated.

Fisher's exact test addresses the same general question without relying on the large-sample chi-square approximation, making it useful when expected counts are small.

## Association between Treatment and clinical response

The following table compares treatment assignment with clinical response at 12 weeks.

```{r cat-response-table}
resp_tab <- table(Treatment=dat$treatment_arm,
                  Response=dat$clinical_response_12w)
resp_tab
```

Under the null hypothesis,

\[
H_0:\text{clinical response is independent of treatment}.
\]

Before selecting the test, examine the expected counts.

```{r cat-response-expected}
resp_chi <- suppressWarnings(chisq.test(resp_tab, correct=FALSE))
round(resp_chi$expected,2)
```

Expected counts are calculated under the assumption that treatment and clinical response are unrelated. They determine whether the chi-square approximation is sufficiently reliable.

## Automated chi-square or Fisher selection

The following block checks the expected counts and performs the appropriate test.

```{r cat-response-select}
expected <- resp_chi$expected
chi_ok <- all(expected >= 5)
# OR
chi_ok <- all(expected>=1) && mean(expected<5)<=.20

if(chi_ok){
  selected_test <- "Pearson chi-square test"
  fit <- resp_chi
} else {
  selected_test <- "Fisher exact test"
  fit <- fisher.test(resp_tab)
}

tibble(
  test=selected_test,
  minimum_expected=min(expected),
  percent_expected_below_5=100*mean(expected<5),
  p_value=fit$p.value
) |> kbl(digits=3, caption="Selected test for treatment and clinical response")
```

The decision is straightforward:

- if expected counts are adequate, use **Pearson's chi-square test**;
- if expected counts are sparse, use **Fisher's exact test**.

A small p-value indicates evidence that clinical response differs across treatment groups. The global test does not show **which treatment differs from Control** or **how large the difference is**, so effect estimates should also be reported.

## Risk differences and risk ratios

For a binary treatment outcome, risks and their contrasts provide a direct measure of treatment effect.

```{r cat-response-effects}
resp_effects <- bind_rows(
  risk_contrast(dat,"clinical_response_12w","Drug_A"),
  risk_contrast(dat,"clinical_response_12w","Drug_B")
) |> mutate(across(c(risk_exposed,risk_reference,risk_difference,rd_low,rd_high), ~100*.x))

kbl(resp_effects, digits=2,
    caption="Clinical-response effects versus Control; risks and risk differences are percentage points")
```

The **risk difference** measures the absolute difference in response probability:

\[
RD=P(Y=1\mid\text{treatment})-P(Y=1\mid\text{Control}),
\]

whereas the **risk ratio** compares the two probabilities proportionally:

\[
RR=\frac{P(Y=1\mid\text{treatment})}
         {P(Y=1\mid\text{Control})}.
\]

The global chi-square or Fisher test provides evidence of association; the risk difference and risk ratio describe the **size of the treatment effect**.

# Fisher’s exact test for sparse contingency tables

Each active treatment is compared separately with Control.

```{r cat-ae-fisher}
tab_A <- with(droplevels(subset(dat,treatment_arm %in% c("Control","Drug_A"))),
              table(Treatment=treatment_arm,Adverse_event=adverse_event_12w))
tab_B <- with(droplevels(subset(dat,treatment_arm %in% c("Control","Drug_B"))),
              table(Treatment=treatment_arm,Adverse_event=adverse_event_12w))

A <- as.data.frame.matrix(tab_A) |> tibble::rownames_to_column("Treatment")
B <- as.data.frame.matrix(tab_B) |> tibble::rownames_to_column("Treatment")
names(A) <- names(B) <- c("Treatment","No","Yes")

knitr::kables(list(
  kbl(A, caption="Drug_A versus Control"),
  kbl(B, caption="Drug_B versus Control")
))
```

```{r fisher-test}
ft_A <- fisher.test(tab_A)
ft_B <- fisher.test(tab_B)

tibble(
  comparison=c("Drug_A vs Control","Drug_B vs Control"),
  odds_ratio=c(unname(ft_A$estimate),unname(ft_B$estimate)),
  p_value=c(ft_A$p.value,ft_B$p.value),
  p_holm=p.adjust(c(ft_A$p.value,ft_B$p.value),method="holm")
) |> kbl(digits=3, caption="Fisher exact test results")
```

Fisher's exact test is performed separately for Drug_A versus Control and Drug_B versus Control. Holm adjustment accounts for the two comparisons.


# McNemar's test for paired binary data

Chi-square and Fisher tests assume that observations contributing to the contingency table are independent. That assumption does not hold when the **same participants are measured twice**.

For paired binary data, McNemar's test evaluates whether the probability of changing in one direction differs from the probability of changing in the opposite direction.

## Example: SBP category before and after treatment

Classify SBP as below 130 mmHg or at least 130 mmHg at baseline and week 12.

```{r cat-mcnemar-table}
mcn_dat <- dat |> drop_na(sbp_baseline_mmHg,sbp_week12_mmHg) |>
  mutate(
    Baseline=if_else(sbp_baseline_mmHg<130,"<130",">=130"),
    Week12=if_else(sbp_week12_mmHg<130,"<130",">=130")
  )

mcn_tab <- table(Baseline=mcn_dat$Baseline,Week12=mcn_dat$Week12)

kbl(as.data.frame.matrix(mcn_tab), digits=0,
    caption="Paired SBP category at baseline and week 12")
```


The paired \(2\times2\) table can be written as:

| Baseline SBP | Week 12 < 130 | Week 12 ≥ 130 |
|---|---:|---:|
| **Baseline < 130** | \(a\) | \(b\) |
| **Baseline ≥ 130** | \(c\) | \(d\) |

The diagonal cells contain participants who remained in the same category. The off-diagonal cells contain participants who **changed category**.

- \(b\) is the number changing from the first category to the second: changed from below 130 to at least 130
- \(c\) is the number changing in the opposite direction: changed from at least 130 to below 130

McNemar's statistic without continuity correction is

\[
\chi^2=\frac{(b-c)^2}{b+c}.
\]

Only the discordant pairs, \(b\) and \(c\), contribute to the test.

```{r cat-mcnemar-test}
mcnemar.test(mcn_tab, correct=FALSE)

b <- mcn_tab[1,2]
c <- mcn_tab[2,1]

if(b+c>0) binom.test(b,b+c,p=.5)
```

McNemar's test evaluates

\[
H_0:P(\text{change in one direction})
   =P(\text{change in the opposite direction}).
\]

When the number of discordant pairs is small, the exact binomial calculation provides an exact alternative.

## Selecting the method

The analysis can be summarized by three decisions:

| Question | Choice |
|---|---|
| Are the categorical observations independent? | chi-square or Fisher |
| Are expected counts adequate? | chi-square if yes; Fisher if no |
| Are binary measurements paired within participants? | McNemar |
| Are several related hypotheses being tested? | consider multiplicity adjustment |
