# MANOVA for multivariate continuous outcomes

## MANOVA: joint treatment effect across multiple outcomes

MANOVA extends ANOVA to several continuous outcomes analyzed together. Here, it tests whether treatment has an overall effect across SBP change, QoL change, and biomarker change.

```{r manova-data}
man_dat <- dat |> drop_na(sbp_change,qol_change,biomarker_change)

man_fit <- manova(cbind(sbp_change,qol_change,biomarker_change)~treatment_arm,data=man_dat)
summary(man_fit,test="Pillai")
```

A significant MANOVA result indicates that treatment groups differ **across the outcomes jointly**, but it does not identify which outcomes differ. Outcome-specific analyses are therefore used as follow-up tests, with attention to multiple testing.

```{r manova-followup}
summary.aov(man_fit)
```

**MANOVA does not replace the need to define primary and secondary outcomes.** In confirmatory studies, the outcome hierarchy and multiple-testing strategy should be specified in advance.



# Correlation analysis

Correlation measures the strength and direction of association between two continuous variables.

| Method | Measures | Useful when |
|---|---|---|
| **Pearson** | linear association | the relationship is approximately linear without influential outliers |
| **Spearman** | monotonic rank association | data are skewed, contain outliers, or the relationship is not strictly linear |

## Pearson correlation

Pearson's correlation coefficient measures the strength and direction of a **linear relationship**:

\[
r=\frac{\sum_i(x_i-\bar{x})(y_i-\bar{y})}
{\sqrt{\sum_i(x_i-\bar{x})^2\sum_i(y_i-\bar{y})^2}}.
\]

It ranges from \(-1\) to \(1\). Values closer to \(-1\) or \(1\) indicate stronger linear association.

```{r pearson-biomarker}
cor.test(~biomarker_baseline_ng_mL+biomarker_week12_ng_mL,
         data=dat,method="pearson")
```

A scatterplot should accompany Pearson correlation to assess the form of the relationship and identify influential observations.

```{r pearson-biomarker-plot, fig.cap="Baseline and week-12 biomarker with a linear fit."}
ggplot(dat,aes(biomarker_baseline_ng_mL,biomarker_week12_ng_mL,colour=treatment_arm)) +
  geom_point(alpha=.45,na.rm=TRUE) +
  geom_smooth(method="lm",se=FALSE,na.rm=TRUE) +
  scale_colour_manual(values=pal) +
  labs(x="Baseline biomarker (ng/mL)",y="Week-12 biomarker (ng/mL)",colour="Treatment")
```

## Spearman rank correlation

Spearman correlation uses the **ranks of the observations** rather than their original values. It measures the strength and direction of a **monotonic relationship** and is useful for skewed data.

Because CRP is strongly right-skewed, Spearman correlation is appropriate for describing the association between baseline and week-12 CRP.

```{r spearman-crp}
cor.test(~crp_baseline_mg_L+crp_week12_mg_L,
         data=dat,method="spearman",exact=FALSE)
```

A strong Pearson or Spearman correlation does not imply that two measurements **agree**, nor does it establish **causation**.

## Correlation matrix

A correlation matrix summarizes associations among several continuous variables. Here, Spearman correlations are used because some variables are skewed.

```{r correlation-matrix}
cor_vars <- dat |> transmute(
  Age=age_years,BMI=bmi_kg_m2,`Baseline SBP`=sbp_baseline_mmHg,
  HbA1c=hba1c_pct,LDL=ldl_mg_dL,HDL=hdl_mg_dL,eGFR=egfr_mL_min_1_73m2,
  `Baseline CRP`=crp_baseline_mg_L,
  `Baseline biomarker`=biomarker_baseline_ng_mL,
  `Baseline QoL`=qol_baseline_0_100)

cm <- cor(cor_vars,use="pairwise.complete.obs",method="spearman")
round(cm,2)
```

```{r correlation-heatmap, fig.height=6, fig.cap="Spearman correlation matrix for selected baseline variables."}
cm_long <- as.data.frame(cm) |>
  tibble::rownames_to_column("x") |>
  pivot_longer(-x,names_to="y",values_to="rho")

ggplot(cm_long,aes(x,y,fill=rho)) +
  geom_tile(colour="white",linewidth=.4) +
  geom_text(aes(label=sprintf("%.2f",rho)),size=3) +
  scale_fill_gradient2(low="#B45252",mid="white",high="#3F7F8D",limits=c(-1,1)) +
  coord_equal() +
  labs(x=NULL,y=NULL,fill="Spearman ρ") +
  theme(axis.text.x=element_text(angle=45,hjust=1))
```

## Correlation is not causation

Correlation shows whether two variables move together, but it does not show that one causes the other.

A strong correlation between baseline and follow-up measurements may indicate that baseline is useful to include in a regression model. However, the treatment effect is estimated by the regression model, not by the correlation coefficient.



