# Linear regression and diagnostics

## Linear-model framework

Linear regression provides a common language for simple regression, multiple regression, ANOVA, ANCOVA, and interactions. The word *linear* refers to linearity in the **coefficients**, not necessarily a straight-line relationship in the raw predictor. A model containing \(x^2\), \(\log x\), or a spline basis remains a linear regression model if the coefficients enter linearly.

For observation \(i\), a multiple linear regression model is

\[
Y_i=\beta_0+\beta_1X_{i1}+\cdots+\beta_qX_{iq}+\varepsilon_i,
\]

where \(Y_i\) is the outcome, \(X_{i1},\ldots,X_{iq}\) are predictors, the \(\beta\)'s are regression coefficients, and \(\varepsilon_i\) is the unexplained error.

Least squares chooses the coefficients that minimize the **sum of squared errors**:

\[
\text{SSE}=\sum_i(Y_i-\hat Y_i)^2.
\]

The total variation in the outcome is

\[
\text{SST}=\text{SSR}+\text{SSE},
\]

where:

- **SST** = total variation;
- **SSR** = variation explained by the model;
- **SSE** = unexplained variation.

Therefore,

\[
R^2=\frac{\text{SSR}}{\text{SST}},
\]

which is the proportion of outcome variation explained by the model.
Adjusted \(R^2\) also accounts for the number of predictors and therefore discourages adding variables that contribute little.


#### Simple linear regression model

```{r simple-sbp-model}
sbp_simple <- lm(sbp_week12_mmHg~sbp_baseline_mmHg, data=dat)

broom::tidy(sbp_simple, conf.int=TRUE) |>
       kbl(digits=3, caption="Simple regression of week-12 SBP on baseline SBP")

summary(sbp_simple)$r.squared
```

### Visualization

```{r simple-sbp-plot, fig.cap="Week-12 SBP versus baseline SBP with the simple regression line."}
ggplot(dat, aes(sbp_baseline_mmHg, sbp_week12_mmHg)) + 
  geom_point(alpha=.42, na.rm=TRUE) + 
  geom_smooth(method="lm", se=TRUE, na.rm=TRUE) +
  labs(x="Baseline SBP (mmHg)", y="Week-12 SBP (mmHg)")
```

The slope is the expected difference in week-12 SBP associated with a 1-mmHg higher baseline SBP. In this descriptive model it is not a causal effect of baseline SBP.

## Multiple linear regression

### Prespecified adjustment set

A treatment-adjusted SBP model can include prespecified baseline covariates. The following model uses treatment, age, and BMI in addition to baseline SBP; all are measured before or at randomization.

#### Multiple linear regression specification

```{r multiple-sbp-model}
sbp_multi_dat <- dat |> drop_na(sbp_week12_mmHg, sbp_baseline_mmHg, treatment_arm, age_years, bmi_kg_m2)

sbp_multi <- lm(sbp_week12_mmHg~sbp_baseline_mmHg+treatment_arm+age_years+bmi_kg_m2, data=sbp_multi_dat)

tibble(R2=summary(sbp_multi)$r.squared, adjusted_R2=summary(sbp_multi)$adj.r.squared) |> kbl(digits=3)

broom::tidy(sbp_multi, conf.int=TRUE) |>
       kbl(digits=3, caption="Multiple linear regression for week-12 SBP")

```

A treatment coefficient is the **mean difference between treatment groups after accounting for the other variables in the model**.

If the model includes a **treatment interaction**, the treatment effect depends on the interacting variable and should not be interpreted as one overall treatment effect.


### Multicollinearity

Variance inflation factor (VIF) measures how much collinearity with other predictors increases the variance of a regression coefficient. Values above 5 may indicate problematic collinearity, while values above 10 are often considered serious. These are guidelines, not absolute cutoffs.

#### Collinearity assessment

```{r multiple-sbp-vif}
car::vif(sbp_multi)
```

## Model assumptions and diagnostics

Linear regression assumes that the model describes the outcome appropriately and that the residuals behave reasonably.

The main checks are:

- **Linearity:** residuals should not show a clear curved pattern.
- **Constant variance:** residual spread should be roughly similar across fitted values.
- **Normality:** the residual Q–Q plot should not show major departures from normality.
- **Influential observations:** no single observation should strongly affect the fitted model.
- **Independence:** observations should not be correlated unless that dependence is modeled.

### Residual diagnostics

```{r multiple-sbp-diagnostics, fig.width=9, fig.height=3.3, fig.cap="Residual-versus-fitted, Q–Q, and Cook's-distance diagnostics for the multiple SBP model."}
par(mfrow=c(1,3)); plot(sbp_multi,which=c(1,2,4)); par(mfrow=c(1,1))
```

### Test for unequal residual variance

```{r multiple-sbp-bp}
lmtest::bptest(sbp_multi)
```

A small Breusch–Pagan p-value suggests that the residual variance is not constant.

### Robust standard errors

If the mean model is appropriate but the residual variance is unequal, **HC3 robust standard errors** can provide more reliable confidence intervals and p-values.

```{r multiple-sbp-robust}
robust_lm_tidy(sbp_multi) |> 
  kbl(digits=3,caption="HC3-robust coefficient inference for the SBP model")
```

