# ANCOVA and interactions for continuous outcomes

## Why adjust for baseline?

When a continuous outcome is measured at baseline and follow-up, ANCOVA compares treatment groups at follow-up while adjusting for the baseline measurement. This can improve precision compared with analyzing change scores alone [@vickers2001].

A basic ANCOVA model is

\[
Y_{12}=\beta_0+\beta_1Y_0+\beta_2I(A)+\beta_3I(B)+\varepsilon.
\]

Here, \(\beta_2\) and \(\beta_3\) represent adjusted treatment differences relative to Control.

Before interpreting these as constant treatment effects, check whether the treatment effect **changes with the baseline value** by adding a baseline-by-treatment interaction.

```{r ancova-hc3-interaction-helper, include=FALSE}
hc3_interaction_test <- function(model){
  b <- coef(model); idx <- grep(":",names(b),fixed=TRUE)
  L <- matrix(0,length(idx),length(b),dimnames=list(names(b)[idx],names(b)))
  L[cbind(seq_along(idx),idx)] <- 1
  car::linearHypothesis(model,L,rhs=rep(0,length(idx)),
                        vcov.=sandwich::vcovHC(model,type="HC3"),test="F")
}
```

# ANCOVA with baseline-by-treatment interaction

The biomarker provides an example where one treatment effect is not appropriate for all baseline values. 

## Analysis dataset

```{r biomarker-model-data}
bio_dat <- dat |>
  dplyr::select(treatment_arm,biomarker_baseline_ng_mL,biomarker_week12_ng_mL) |>
  drop_na()

bio_center <- median(bio_dat$biomarker_baseline_ng_mL)
bio_dat <- bio_dat |> mutate(base_c=biomarker_baseline_ng_mL-bio_center)
```

Because regression coefficients are interpreted when the other predictors equal **zero**, a raw baseline value of zero may give an unrealistic treatment comparison. Therefore, baseline is centered at its median, so that a centered value of zero corresponds to the **median baseline biomarker value** and the treatment coefficients represent group differences at that typical baseline level.

## Test for interaction

```{r biomarker-interaction-test}
bio_common <- lm(biomarker_week12_ng_mL~base_c+treatment_arm,data=bio_dat)
bio_int <- lm(biomarker_week12_ng_mL~base_c*treatment_arm,data=bio_dat)

hc3_interaction_test(bio_int)
```

The interaction tests whether the **treatment effect changes with baseline biomarker**. Here, the interaction is important, so the interaction model is retained.

## Treatment effects at different baseline values

Because the treatment effect changes with baseline biomarker, **one overall treatment effect would be misleading**. Instead, treatment groups are compared at representative baseline values: the 25th, 50th, and 75th percentiles.

```{r biomarker-emmeans}
bio_q <- quantile(bio_dat$biomarker_baseline_ng_mL,c(.25,.50,.75))
bio_at <- as.numeric(bio_q-bio_center)
bio_levels <- tibble(base_c=bio_at,quantile=c("25th percentile","50th percentile","75th percentile"))

bio_emm <- emmeans::emmeans(
  bio_int,~treatment_arm|base_c,
  at=list(base_c=bio_at),
  vcov.=sandwich::vcovHC(bio_int,type="HC3")
)

summary(pairs(bio_emm,adjust="tukey"),infer=c(TRUE,TRUE)) |>
  left_join(bio_levels,by="base_c") |>
  mutate(baseline_biomarker=base_c+bio_center) |>
  dplyr::select(quantile,baseline_biomarker,contrast,estimate,lower.CL,upper.CL,p.value) |>
  kbl(digits=3,caption="Treatment effects at the 25th, 50th, and 75th percentiles of baseline biomarker")
```

A negative active-treatment versus Control difference means a lower adjusted week-12 biomarker at that baseline value.

## Interaction plot

```{r biomarker-interaction-plot, fig.cap="Observed biomarker values and fitted treatment-specific relationships."}
bio_grid <- expand_grid(
  base_c=seq(min(bio_dat$base_c),max(bio_dat$base_c),length.out=150),
  treatment_arm=levels(bio_dat$treatment_arm)
)

bio_grid$predicted <- predict(bio_int,newdata=bio_grid)
bio_grid$baseline_biomarker <- bio_grid$base_c+bio_center

ggplot(bio_dat,aes(biomarker_baseline_ng_mL,biomarker_week12_ng_mL,colour=treatment_arm)) +
  geom_point(alpha=.24) +
  geom_line(data=bio_grid,aes(baseline_biomarker,predicted,colour=treatment_arm),linewidth=1.05) +
  scale_colour_manual(values=pal) +
  labs(x="Baseline biomarker (ng/mL)",
       y="Week-12 biomarker (ng/mL)",
       colour="Treatment")
```

Different slopes show why the treatment effect must be interpreted **at specific baseline biomarker values**, rather than as one overall mean difference. :contentReference[oaicite:4]{index=4}

## Diagnostics

```{r biomarker-interaction-diagnostics, fig.width=9, fig.height=3.5}
par(mfrow=c(1,3)); plot(bio_int,which=c(1,2,4)); par(mfrow=c(1,1))
```

## Key takeaway
The analysis is based on a **linear regression model** fitted with `lm()` that includes a baseline-by-treatment interaction. The interaction is tested with `hc3_interaction_test()`, and treatment effects at selected baseline values are estimated using `emmeans()` and `pairs()`. **HC3 robust standard errors**, calculated with `sandwich::vcovHC(..., type="HC3")`, are used for both tests and treatment comparisons.


# Log-scale ANCOVA

CRP is positive and strongly right-skewed, so a log-scale model is more appropriate than modeling raw CRP directly. 

\[
\log(CRP_{12})
=
\beta_0+\beta_1\log(CRP_0)+\beta_2I(A)+\beta_3I(B)+\varepsilon.
\]

```{r crp-log-data}
crp_anc <- dat |>
  filter(!is.na(crp_baseline_mg_L),!is.na(crp_week12_mg_L),
         crp_baseline_mg_L>0,crp_week12_mg_L>0)

crp_common <- lm(log(crp_week12_mg_L)~ log(crp_baseline_mg_L)+treatment_arm,data=crp_anc)

crp_int <- lm(log(crp_week12_mg_L)~ log(crp_baseline_mg_L)*treatment_arm,data=crp_anc)

hc3_interaction_test(crp_int)
```

The interaction is not needed, so the common-slope log model is retained.

## Treatment effects as geometric-mean ratios

Because CRP is modeled on the **log scale**, `emmeans()` estimates baseline-adjusted treatment means on that scale. `contrast()` then compares each active treatment with Control.

```{r crp-log-contrasts}
crp_emm <- emmeans::emmeans(
  crp_common,~treatment_arm,
  vcov.=sandwich::vcovHC(crp_common,type="HC3")
)

crp_ctrl <- summary(
  contrast(crp_emm,method="trt.vs.ctrl",ref=1,adjust="holm"),
  infer=c(TRUE,TRUE)
) |>
  mutate(ratio=exp(estimate),
         ratio_low=exp(lower.CL),
         ratio_high=exp(upper.CL)) |>
  dplyr::select(contrast,ratio,ratio_low,ratio_high,p.value)

kbl(crp_ctrl,digits=3,
    caption="Adjusted CRP geometric-mean ratios")
```

The treatment contrasts are first estimated as differences on the **log scale**. Exponentiating these differences converts them to **ratios of adjusted geometric means**:

\[
\text{Ratio}
=
\frac{\text{adjusted geometric mean in treatment}}
{\text{adjusted geometric mean in Control}}.
\]

A ratio of **1** indicates no difference between groups. A ratio of **0.80** means the active-treatment group has about **20% lower CRP**, while a ratio of **1.25** means about **25% higher CRP**, after adjustment for baseline CRP.


## Diagnostics

```{r crp-log-diagnostics, fig.width=8.5, fig.height=3.8}
par(mfrow=c(1,2)); plot(crp_common,which=1); plot(crp_common,which=2); par(mfrow=c(1,1))
```

## Key interpretation

The meaning of a treatment effect depends on the model:

- **Standard ANCOVA:** adjusted mean difference.
- **Interaction model:** treatment effect depends on baseline value.
- **Log-scale ANCOVA:** exponentiated treatment effect is a ratio of adjusted geometric means.

Always interpret the treatment effect according to the model that was fitted.

