# Survival analysis and time-to-event models

Survival analysis is used when the outcome is **time until an event**, while allowing for participants who do not experience the event during follow-up.

## Time-to-event data and censoring

A participant who remains event-free at the end of follow-up is **censored**. The exact event time is unknown, but the participant is known to have remained event-free up to the last observed time.

For example, if a participant has no event by day 90,

\[
T>90.
\]

```{r survival-data}
surv_dat <- dat |>
  dplyr::select(treatment_arm,time_to_event_days,event_observed_90d) |>
  drop_na()
```

```{r survival-summary}
surv_dat |>
  group_by(treatment_arm) |>
  summarise(n=n(),
            events=sum(event_observed_90d==1),
            censored=sum(event_observed_90d==0),
            .groups="drop") |>
  kbl(digits=0,caption="Events and censoring by treatment group")
```

The precision of survival analyses depends strongly on the **number of observed events**.

## Kaplan–Meier curves and log-rank test

The Kaplan–Meier method estimates the probability of remaining event-free beyond time \(t\):

\[
S(t)=P(T>t).
\]

```{r km-fit}
surv_obj <- survival::Surv(surv_dat$time_to_event_days,
                           surv_dat$event_observed_90d)

km_fit <- survival::survfit(surv_obj~treatment_arm,data=surv_dat)

summary(km_fit,times=c(30,60,90))
```

```{r km-plot, fig.cap="Kaplan–Meier event-free survival by treatment group."}
plot(km_fit,col=unname(pal),lwd=2,mark.time=TRUE,
     xlab="Days",ylab="Event-free survival",conf.int=FALSE)

legend("bottomleft",legend=levels(surv_dat$treatment_arm),
       col=unname(pal),lwd=2,bty="n")
```

The **log-rank test** provides a global comparison of the survival curves.

```{r logrank}
survival::survdiff(surv_obj~treatment_arm,data=surv_dat)
```

A small p-value indicates evidence that survival differs across treatment groups.

## Cox proportional-hazards regression

The Cox model estimates how treatment or other predictors are associated with the **hazard of experiencing the event**:

\[
h(t\mid X)=h_0(t)\exp(\beta^TX),
\]

where \(h_0(t)\) is the baseline hazard.

Exponentiating a coefficient gives a **hazard ratio**:

\[
HR=e^\beta.
\]

A hazard ratio below 1 indicates a lower hazard, while a value above 1 indicates a higher hazard relative to the reference group.

```{r cox-models}
cox_trt <- survival::coxph(
  survival::Surv(time_to_event_days,event_observed_90d)~treatment_arm,
  data=surv_dat,ties="efron"
)

broom::tidy(cox_trt,exponentiate=TRUE,conf.int=TRUE) |>
  kbl(digits=3,caption="Treatment Cox proportional-hazards model")
```

A hazard ratio compares the **instantaneous hazard among participants who are still at risk**. It should not be interpreted as a risk ratio at a fixed time point.

## Proportional-hazards assumption

The Cox model assumes that the hazard ratio is approximately **constant over time**.

```{r cox-ph}
cox_ph <- survival::cox.zph(cox_trt)
cox_ph
```

```{r cox-ph-plot, fig.height=4.5, fig.cap="Schoenfeld-residual diagnostic for the Cox model."}
plot(cox_ph)
```

A small p-value or a clear time-related pattern suggests that the proportional-hazards assumption may not hold. If so, the appropriate alternative depends on how the treatment effect changes over time and on the scientific question.

## Key distinction

These models describe different types of outcomes:

| Model | Outcome | Effect measure |
|---|---|---|
| Logistic regression | whether an event occurred | odds ratio |
| Poisson or negative binomial regression | number of events | count or rate ratio |
| Cox regression | time until an event | hazard ratio |

The model should be chosen according to the **outcome and the effect of interest**.

