## Power and sample size

Power calculations are mainly used **before a study begins** to decide how many participants are needed.

**Power** is the probability of detecting a real effect of a specified size. For example, 90% power means that if the true effect is the one assumed in the calculation, the study has about a 90% chance of producing a statistically significant result at the chosen significance level.

Sample size depends on several assumptions:

- the smallest clinically important effect;
- expected variability for continuous outcomes;
- expected event or response rates for binary outcomes;
- significance level, usually \(\alpha=0.05\);
- desired power, commonly 80% or 90%;
- allocation ratio between groups;
- expected dropout or missing data.

Larger samples are usually required when the expected treatment effect is small, variability is high, or higher power is required.

### Continuous-outcome planning example

Suppose a two-arm trial is designed to detect a **5-mmHg difference in mean SBP**, with an expected SD of **12 mmHg**, a two-sided significance level of 0.05, and 90% power.

```{r power-continuous}
power.t.test(
  delta=5,
  sd=12,
  sig.level=.05,
  power=.90,
  type="two.sample",
  alternative="two.sided"
)
```

Here:

- `delta=5` is the mean difference we want to detect;
- `sd=12` is the expected within-group standard deviation;
- `sig.level=.05` is the Type I error rate;
- `power=.90` requests 90% power.

The function returns the approximate **sample size required per group**.

### Binary-outcome planning example

Suppose the expected clinical response rate is **30% in the control group** and **45% in the treatment group**.

```{r power-binary}
power.prop.test(
  p1=.30,
  p2=.45,
  sig.level=.05,
  power=.90,
  alternative="two.sided"
)
```

Here, the effect of interest is the difference between response rates:

\[
0.45-0.30=0.15.
\]

The calculation estimates how many participants are needed to detect this **15-percentage-point difference** with 90% power.

These calculations are for **planning**. After the study is completed, the estimated treatment effect and its confidence interval are more informative than calculating "observed power" from the observed result.

## Standard deviation, standard error, and confidence intervals

These quantities describe different things.

**Standard deviation (SD)** describes how much individual observations vary.

For example, if mean SBP is 130 mmHg with SD 12 mmHg, the SD describes the spread of individual SBP measurements around the mean.

**Standard error (SE)** describes the uncertainty in an estimated statistic, such as the sample mean.

For a sample mean,

\[
SE(\bar{x})=\frac{s}{\sqrt{n}}.
\]

For example, if \(s=12\) and \(n=100\),

\[
SE=\frac{12}{\sqrt{100}}=1.2.
\]

Thus, individual SBP values may vary considerably (\(SD=12\)), while the estimated mean can still be relatively precise (\(SE=1.2\)).

A **confidence interval (CI)** gives a range of plausible values for the population parameter. An approximate 95% CI for a mean is

\[
\bar{x}\pm1.96\,SE.
\]

If the estimated mean is 130 and \(SE=1.2\),

\[
130\pm1.96(1.2)\approx(127.6,\;132.4).
\]

In short:

**SD = variability between individuals**

**SE = uncertainty in the estimate**

**CI = plausible range for the population parameter**

Increasing sample size usually decreases the SE and narrows the confidence interval, but a large sample cannot correct bias caused by poor study design, measurement error, or inappropriate analysis.